---
title: "dplyr, plyr, and tidyr: Baby Names in the US"
author: "Sam Castillo"
date: "April 15, 2017"
output: html_document
fontsize: 11pt
---

```{r global_options, warning = F, message = F}
library(plyr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(gridExtra)
library(googleVis)
library(reshape2)
```

A dataset of baby names used by Hadley Wickham for a workshop can be found at https://github.com/hadley/babynames/tree/master/data.
It contains the 1000 most popular male and female baby names in the US, from 1880 to 2008. There
are 258,000 records (1000 * 2 * 129) but only four variables: year, name, sex, and percent. 

The task is to identify the top 5 boys and girls names for each year from 1880 to 2008 and put it into a dataframe.

```{r}
bnames = read.csv('bnames.csv')
str(bnames)
#View(bnames)
#bnames = as_data_frame(bnames)
summary(bnames)
```

We can arrange by percent and by year with dplyr.

```{r}

tmp = bnames

boys = tmp %>%
  filter(sex == "boy")%>%
  group_by(name, year) %>%
  summarise(sum_percent = sum(percent))%>%
  select(-year) %>%
  arrange(desc(sum_percent))

head(boys,10)

tmp = bnames

girls = tmp %>%
  filter(sex == "girl")%>%
  group_by(name, year) %>%
  summarise(sum_percent = sum(percent))%>%
  select(-year) %>%
  arrange(desc(sum_percent))

head(girls,10)
```

Finding the top 5 for each year and putting it in a wider format is more difficult.  The function `spread` from tidyr would not work easily in this case, at least not from what I can tell.  Here I split the dataframe into a list of dataframes and then apply a function over the entire list.  Finally, I put everything back together.

```{r}
boys_tmp = subset(bnames, sex =="boy")
boys_tmp = split(boys_tmp, boys_tmp$year)

girls_tmp = subset(bnames, sex =="girl")
girls_tmp = split(girls_tmp, girls_tmp$year)

```

```{r, inculde = T, echo = T}
top5 = function(dat){
  cur_year = dat$year[1]
  top_rows = top_n(dat,5, percent)
  out = c(cur_year, as.character(top_rows$name))
  # df = as_data_frame(matrix(ncol = 6, nrow = 1))
  # df[1,] = out
  return(out)
}
```

Applying over the whole dataset.

```{r}
mynames = c("year", "1st.name" ,"2nd.name" , "3rd.name", "4th.name", "5th.name")

girls_final = NULL
girls_final = sapply(girls_tmp, top5)
boys_final = ldply(boys_tmp, top5)

df_girls_final = setNames(do.call(rbind.data.frame, girls_final), mynames)

df_girls_final = df_girls_final[1:6] %>% as_data_frame()

df_boys_final = boys_final %>%
  select(-1) 

names(df_boys_final) = mynames
  
df_final = rbind(df_girls_final, df_boys_final)

head(df_final, 5)
tail(df_final, 5)
```

We can try to identify the "trendiest" baby names by fitting linear regression models. We only use simple linear regression here.

```{r}

data <- bnames

#creates a function of temp
lm.fit <- function(temp){
  #fits a simple linear regression model with year as the predictor and percent as the response
  #over the columns percent and year of the data temp which was input
  fit <- lm( percent ~ year, data = temp)
  #returns the intercept and slope of the regression line, and n, the number of rows
  return(data.frame(int=fit$coef[1],slope=fit$coef[2],
                    
  n=dim(temp)[1]))
}
#For each boys name and girls name, apply the lm.fit function to return a row for each name.
#the columns are name, sex, intercept, slope, and n respectively.  See the output from the head 
#function below
inc.dec <- ddply(data,.(name,sex),lm.fit)

#Examine only those names with greater than 100 observations
inc.dec <- subset(inc.dec,n>100)

#subset again to only the most extreme cases.  This looks only at the top 1% and botton 1% of the intercepts.
#the na.rm options removes missing values.  
inc.dec <- subset(inc.dec, (slope > quantile(slope, p=0.99,na.rm=T))|(slope < quantile(slope, p=0.01, na.rm=T)))
head(inc.dec)
dim(inc.dec)
```

The data.frame inc.dec above has 16 rows. For each of those names, I make a scatterplot with year on
the x-axis and percent on the y-axis. Then I label the plot with the name, and use abline() to add the least
squares regression line. 

```{r eval = T, echo = T}
outliers_df = filter(bnames, name %in% inc.dec$name)

ggplot_function = function(cur_name) {
  cur_dat = subset(outliers_df, name == cur_name)# %>%select(-sex, -name)
  cur_dat = cur_dat[cur_dat$percent > 0.001,] #Remove rows with percent equal to zero
  names(cur_dat) = c("year", "name", "percent", "sex")
  lm.dat = subset(inc.dec, name == cur_name)
  p1 = ggplot(data = cur_dat, aes(year,percent)) + geom_point(size = 0.2, colour = "red") + geom_abline(data = lm.dat, slope = lm.dat$slope, intercept = lm.dat$int) + theme_light() + ggtitle(paste(cur_name)) + xlim(1880, 2008) + ylim(0, 0.06)
  
  p1
}
```

```{r }
plots = lapply(inc.dec$name, ggplot_function)
```


```{r}
do.call("grid.arrange", plots)
```

Have most babies had similar names in certain years?  In other words, how has the sum of percentage of the top 100 baby names changed over time?

Here I create a plot that shows (by year and gender) the proportion of US children who have a name in the
top 100. Proportion is on the y-axis, year on the x-axis, and two lines, one for each
gender.

```{r results = 'asis'}
suppressPackageStartupMessages(library(googleVis))

df_boys = bnames %>%
  filter(sex =="boy") %>%
  group_by(year) %>%
  arrange(desc(percent)) %>%
  filter(percent > min(head(percent, 101))) %>%
  group_by(year)%>%
  mutate(sum_percent = sum(percent)) 
 
df_girls = bnames %>%
  filter(sex =="girl") %>%
  group_by(year) %>%
  arrange(desc(percent)) %>%
  filter(percent > min(head(percent, 101))) %>%
  group_by(year)%>%
  mutate(sum_percent = sum(percent)) 
head(df_girls)

p2 = ggplot(data = rbind(df_boys, df_girls), aes(year, sum_percent, sex))

p2 + geom_point(aes(color = factor(sex))) + ggtitle("Sum of Top 100 Percentage of Baby Names by Year") + theme_linedraw()

df3 = rbind(df_boys, df_girls)

df3$year = as.Date(as.character(df3$year), "%Y")

Anno = gvisAnnotationChart(df3, 
                           datevar = "year",
                           numvar = "sum_percent",
                           idvar = "sex"
                           )
Anno
```

